### 2013-02-07 doc/xps2013/proposal.tex
research items

\comment{mostly a copy and paste of the pomace?}

@@

@@

>>>>

Grappa applications are likely to use large amounts of data, 100s of terabytes
to petabytes. Keeping all these data in DRAM only is unlikely to be cost
effective. We will explore mechanisms that allow data to reside in solid-state
storage {\em only}.

Grappa is very effective in hiding latencies of network communication. In
fact, our measurements and estimates show that we might be able to tolerate
even more latencies, on the order of milliseconds per request. Therefore, this
suggests that we might be able to keep data in SSD only and service data
access requests directly from that, without leaving copies in DRAM.

The natural point of integration for non-volatile memories in Grappa is to
store the global distributed shared heap in non-volatile memory, since for
irregular applications the heap often holds data that is infrequently
modified. All accesses to this shared heap in Grappa programs are done through
an \textls*{\small API}, so this change could be made transparent to the
programmer.

If an access is to a chunk of the heap stored in a node's local memory, Grappa
issues a prefetch for the address and context switches to other work while the
prefetch is executing~\cite{Nelson:hotpar2011}. If the access is to memory
stored on another node, Grappa issues a message to that node to request the
data and context switches to other work until the reply arrives. Non-volatile
memory would be accessed in the same way: If a request is to memory stored in
non-volatile memory, Grappa would initiate the request and context switch to
other work until the request had completed.

This is not straightforward for three reasons: (1) {\bf \textls*{\small
SSD}\hspace{0.75pt}s assume locality.} In our benchmarks, we find the average
request size to be less than 64 bytes. Today's \textls*{\small
SSD}\hspace{0.75pt}s are optimized for reads in the multi-kilobyte range. (2)
{\bf \textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The
fastest~\cite{fusionio} can perform on the order of 1 million random reads per
second, and many are limited to less than 100K/s. Even the individual flash
chips are slow, on the order of 40K/s~\cite{micronFlash}. In contrast, cores
may be idle in Grappa if each node cannot perform 100 million random reads per
second of the global heap. And (3) {\bf \textls*{\small SSD}\hspace{0.75pt}s
have high overhead.} Mass market \textls*{\small SSD}\hspace{0.75pt}s are
usually treated as normal disks: accesses must occur through the kernel,
incurring a high context-switch overhead. Grappa depends on user-mode access
to the network layer to avoid this overhead; a similar approach has been
applied to \textls*{\small SSD}\hspace{0.75pt}s in recent
research~\cite{caulfield:2012}.

We see several ways to address these challenges: \noindent{\bf Scaling up the
Flash array} Existing Flash-based \textls*{\small SSD}\hspace{0.75pt}s are
designed for large block requests. An \textls*{\small SSD} for Grappa could be
designed to support high concurrency, small requests. Where a traditional
\textls*{\small SSD} might put multiple Flash chips on the same bus, or use
Flash chips in parallel to build a wider, high-bandwidth bus, a Grappa
\textls*{\small SSD} would try to maximize parallelism by putting each chip on
its own bus. Serial Flash might be used to minimize pin count requirements, at
the cost of some additional latency. This design could scale to request rates,
but the challenge would be chip count: to support 100 million reads per
second, an array of 25$\mu$s flash would require 2500 chips. This is possible,
but expensive.

\noindent{\bf Scaling down the processors.} Alternatively, rather than scaling
up the flash array, processors could be scaled down to match the flash request
rate. A low-power processor with few cores might generate only 1 million
references per second. This could be satisfied with only $25\times 25\mu$s
Flash chips. There are two problems with this design for our purpose. First,
while each node requires fewer chips, the overall cluster performance goal,
and thus chip count requirement, is the same. Second, low single-threaded
performance is detrimental to performance during periods of diminished
application concurrency.

\noindent{\bf Buffering requests to increase locality.} Grappa trades latency
for better throughput by aggregating network communications for better
utilization and by migrating synchronization operations for reduced
contention. The same idea could be applied to increase locality and amortize
the cost of Flash accesses. Requests to nearby flash locations would be
bucketed by flash page address for a limited time. When a sufficient number of
requests have arrived for a page or when the requests time out, a single page
request would be issued to the Flash layer; its result would be used to
satisfy the individual requests.

@@

Language and compiler support are instrumental in making Grappa more usable.
We envision language and compiler support for three main aspects: (1) specify
the memory model and semantics of global memory accesses; (2) automatic
insertion of calls to the runtime, to avoid having the end programmer insert
explicit calls; and (3) performing code optimizations specific for Grappa as
well as traditional optimizations in ways that will not break program
semantics.

Specifying the memory model and access semantics of the Grappa runtime is very
important for several reasons. The programmer needs to know the behavior of
accesses to the global heap. The compiler needs to know that too so that code
optimizations can be performed without breaking program semantics. And the
model has to be relaxed enough to enable efficient implementations. We opted
for a sequential consistency for data-race-free programs, similarly to the
C++0x memory model (something programmers and system software designers are already used to).

We will develop type qualifiers to let programmers declare shared data, e.g.,
\texttt{shared int}. This will enable the compiler to automatically insert
runtime calls for accesses to global heap data as well some forms of
synchronized operations (e.g. atomic fetch-and-add) to global variables. This
will also enable new analyses and optimizations, for example, the compiler
could perform spatial and temporal locality analysis to improve request
aggregation when it can proved that several accesses go to the same
destination.

@@

In order to facilitate the development and evaluation of Grappa, we will
develop a comprehensive benchmark suite for graph analytics applications. This
will not only be useful for us, but the whole community at large, since there
is a lack of benchmarks in this application space.

This benchmark suite will consist of application kernels, reference
implementations, data-sets, data-set generators, documentation, specifications
for generating standardized performance results, and a hosting service
(http://graphbench.org) to provide public access to the benchmark and datasets
as well as aggregate and publish results by all users of the benchmark suite.

This benchmark suite will be a tremendous value to the research community. Not
only will it enable researchers to evaluate ideas on graph processing system
techniques, but it will also enable fair comparisons of ideas in a depth not
currently available from other graph related benchmark suites. The benchmark
will be released in a completely open fashion, so anyone will be able to use
it (and contribute to it), there will be no licensing issues even if companies
(such as Oracle) want to use it. The goal is to get something out relatively
early in the hope of (i) getting others to cluster around this work rather
than develop alternatives, and (ii) help establish some of the metrics about
how to present performance results.

We will build a collection of key computational kernels from graph and
irregular applications. The benchmark suite will consist of both documentation
on the actual kernel (ideally a research paper or book reference sourced from
the community), and a reference implementation. The benchmark suite will also include sample data collected from existing sources, and synthetic
workload generators. The collected data provides a grounding for performance
analysis on real-world problems. The workload generators provide cleaner,
easier to reason about randomly generated data-sets at a programmable scale.
